\textbf{Seasonal Dynamics of Epiphytic Microbial Communities on Marine Macrophyte Surfaces}

### 2020-11-21 submission/manuscript.tex
Text formatting.

\sisetup{mode=text}
\setlength\parindent{24pt}

@@

Marine macrophytes (seagrasses and macroalgae) are important ecosystem
engineers forming close associations with microorganism belonging to all
three domains of life (Egan \emph{et al.}, 2013; Tarquinio \emph{et
al.}, 2019). Microbes can live within macrophyte tissue as endophytes or
can form epiphytic communities on surfaces of leaves and thalli (Egan
\emph{et al.}, 2013; Hollants \emph{et al.}, 2013; Tarquinio \emph{et
al.}, 2019). Epiphytic and endophytic microbial communities exhibit a
close functional relationship with the macrophyte host. It was proposed
that this close relationship constitutes a holobiont, an integrated
community where the macrophyte organism and its symbiotic partners
support each other (Margulis, 1991; Egan \emph{et al.}, 2013; Tarquinio
\emph{et al.}, 2019).

@@

The majority of studies describing macrophyte epiphytic microbial
communities did not include possible seasonal changes (Crump and Koch,
2008; Lachnit \emph{et al.}, 2009; Burke and Thomas \emph{et al.}, 2011;
Roth-Schulze \emph{et al.}, 2016; Ugarelli \emph{et al.}, 2019). If
seasonal changes were taken into account, low temporal frequency,
applied methodologies and/or limited number of analyzed host species did
not allow a high taxonomic resolution (Tujula \emph{et al.}, 2010;
Lachnit \emph{et al.}, 2011; Miranda \emph{et al.}, 2013; Michelou
\emph{et al.}, 2013; Mancuso \emph{et al.}, 2016). In the present study
we describe the seasonal dynamics of bacterial and archaeal communities
on the surfaces of the seagrass \emph{Cymodocea nodosa} and siphonous
macroalgae \emph{Caulerpa cylindracea} determined on a mostly monthly
scale. Bacterial and archaeal epiphytes were sampled in a meadow of
\emph{C. nodosa} invaded by the invasive \emph{C. cylindracea} and in a
locality of only \emph{C. cylindracea} located in the proximity of the
seagrass meadow. For comparison, the microbial community of the ambient
seawater was also characterized.

@@

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of sequences without
known relatives (no relative sequences) and eukaryotic, chloroplast and
mitochondrial sequences (\autoref{supp-nseq_notus}). A total of 35
samples originating from epiphytic archaeal and bacterial communities
associated with surfaces of the seagrass \emph{C. nodosa} and the
macroalga \emph{C. cylindracea} were analyzed. In addition, 18 samples
(one of the samples was sequenced twice) originating from the ambient
seawater were also processed for comparison. The number of reads per
sample ranged between 8,408 and 77,463 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off as commonly observed in
high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences as mentioned above reads were clustered into 28,750
different OTUs at a similarity level of 97 \si{\percent}. Read numbers
were normalized to the minimum number of sequences (8,408,
\autoref{supp-nseq_notus}) through rarefaction resulting in 16,986
different OTUs with 361 to 2,026 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes in richness
and diversity the observed number of OTUs, Chao1, ACE, Exponential
Shannon and Inverse Simpson (Jost, 2006) were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{C. cylindracea} (mixed {[}Number of OTUs, 1,678.1 ±
143.6 OTUs{]} and monospecific {[}Number of OTUs, 1,734.2 ± 150.9
OTUs{]}) than for \emph{C. nodosa} (Number of OTUs, 1,054.2 ± 213.0
OTUs) and lowest values were obtained for the microbial community of the
ambient seawater (Number of OTUs, 529.1 ± 136.4 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not reveal such large
dissimilarities. \emph{C. nodosa} communities showed a slow increase
towards the end of the study, while \emph{C. cylindracea} (mixed and
monospecific) communities were characterized by slightly higher values
in spring and summer than in autumn and winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient, respectively, were calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from the
particular environment. The highest proportion of shared OTUs and
community was found between mixed and monospecific \emph{C. cylindracea}
(Jaccard, 0.35; Bray-Curtis, 0.77), while lower shared values were
calculated between ambient seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{C. nodosa} and
\emph{C. cylindracea} were approximately in-between the values of mixed
and monospecific \emph{C. cylindracea}. To assess seasonal changes in
the proportion of shared OTUs and communities the Jaccard's and
Bray-Curtis Similarity Coefficients were calculated between consecutive
sampling points (\autoref{shared}). Both coefficients showed similar
trends. Temporal proportional changes were more pronounced for ambient
seawater than for \emph{C. nodosa} and especially \emph{C. cylindracea}
associated communities (\autoref{shared}). In addition, only 0.4 -- 1.0
\si{\percent} of OTUs from each surface associated community were
present at all seasons. These persistent OTUs constituted a high
proportion of total sequences (36.9 -- 53.0 \si{\percent}). To further
disentangle the environmental and seasonal community dissimilarity a
Principal Coordinates Analysis (PCoA) based on Bray-Curtis distances and
OTU abundances was applied. A clear separation between ambient seawater
and surface associated communities was found (\autoref{pcoa}). In
addition, a separation of epiphytic bacterial and archaeal communities
based on host species was detected. This separation was further
supported by ANOSIM (R = 0.96, \emph{p} \textless{} 0.001). Seasonal
changes of \emph{C. nodosa} associated communities indicated a
separation between spring, summer and autumn/winter samples (ANOSIM, R =
0.53, \emph{p} \textless{} 0.01). For \emph{C. cylindracea} associated
communities a separation between summer and autumn/winter/spring samples
was observed that was, however, not as strong as for \emph{C. nodosa}
associated communities (ANOSIM, R = 0.31, \emph{p} \textless{} 0.01)
(\autoref{pcoa}).

@@

We observed a strong differentiation between the surface attached and
ambient seawater communities at the level of OTUs, in agreement with
most published studies (Burke and Thomas \emph{et al.}, 2011; Michelou
\emph{et al.}, 2013; Roth-Schulze \emph{et al.}, 2016; Mancuso \emph{et
al.}, 2016; Crump \emph{et al.}, 2018; Ugarelli \emph{et al.}, 2019).
This indicates that marine macrophytes are a selecting factor from the
pool of microbial taxa present in the ambient seawater, modifying the
microbial community once the macrophyte associated microbial biofilm
develops (Salaün \emph{et al.}, 2012; Michelou \emph{et al.}, 2013). In
contrast, Fahimipour \emph{et al.} (2017) report in a global study of
\emph{Zostera marina}, similarities between the microbial community
developed on leaves and in the ambient seawater. The discrepancy between
our data and the study of Fahimipour \emph{et al.} (2017) could be
explained by different seagrass species, methodological variations or
biogeographic trends as Fahimipour \emph{et al.} (2017) analyzed samples
from different locations throughout the northern hemisphere. It is
possible that the microbial communities in ambient seawater and on
leaves from the same location are differing but are still more similar
to each other when compared to other sampling locations. Indeed, it was
found that prokaryotic communities vary substantially between different
sampling sites (Bengtsson \emph{et al.}, 2017). When the taxonomic
composition at high ranks was analyzed no such strong differentiation
was noticed. Phyla and classes such as \emph{Actinobacteriota},
\emph{Bacteroidota}, \emph{Cyanobacteria}, \emph{Alphaproteobacteria},
\emph{Gammaproteobacteria} and \emph{Verrucomicrobiota} were found in
both ambient seawater as well as macrophyte associated, in agreement
with previous studies (Burke and Thomas \emph{et al.}, 2011; Egan
\emph{et al.}, 2013; Michelou \emph{et al.}, 2013). In contrast, when
low taxonomic ranks were analyzed (i.e., family and genus) a strong
differentiation was observed (Figs. \ref{cyano} -- \ref{desulfo}). A
similar differentiation at lower microbial taxonomic ranks between
ambient seawater and macrophytes was described for other macrophyte
species as well (Egan \emph{et al.}, 2013; Michelou \emph{et al.}, 2013;
Ugarelli \emph{et al.}, 2019).

@@

Seasonal changes in richness in the epiphytic community were substantial
as indicated by the proportion of OTUs (\(\leq\) 1.0 \si{\percent})
present at every sampling date. These persistent OTUs, however, were
accounting for a high proportion of sequences (\(\leq\) 53.0
\si{\percent}) (\autoref{shared}). A very similar proportion of
persistent OTUs was reported in high-frequency sampling studies
describing seasonal changes in picoplankton (Gilbert \emph{et al.},
2009, 2012). In comparison to the seawater community, a lower degree of
seasonal shifts was observed for the macrophyte surface associated
communities. It appears that microniches at the surfaces of macrophytes
are providing more stable conditions than the ambient seawater. At the
level of OTUs seasonal changes of \emph{C. nodosa} and \emph{C.
cylindracea} associated communities were identified that could be linked
to the growth cycle of the seagrass and macroalgae (M. Najdek, personal
communication). \emph{C. nodosa} was characterized by a spring community
during maximum seagrass proliferation, a summer community during the
highest standing stock of \emph{C.nodosa} and an autumn/winter community
during the decay of seagrass biomass. In contrast, \emph{C. cylindracea}
started to proliferate in late spring and was characterized only by a
summer community during high growth rates and by an autumn/winter/spring
community when the biomass was at the peak and decaying thereafter.
Similar seasonal changes in the epiphytic community were also described
for other macroalgae (Tujula \emph{et al.}, 2010; Lachnit \emph{et al.},
2011). Higher seasonal stability of \emph{C. cylindracea} surface
communities than in \emph{C. nodosa} was also observed in the higher
proportion of shared communities between two consecutive sampling dates
in \emph{C. cylindracea}.

@@

surface for nonspecific colonizers (Zavodnik \emph{et al.}, 1998).

@@

not exhibit strong successional patterns. In every analyzed group, with

@@

and no relative \emph{Desulfocapsaceae}, known sulfate sediment groups

@@

In temperate zones, marine macrophytes are exhibiting growth cycles, so
it is not surprising that the associated epiphytic microbial community
is undergoing partial seasonal changes. In the present study, we could
identify in every analyzed high rank taxa phylogenetic groups present
throughout the year, comprising most of the sequences and a lower
proportion of taxa showing seasonal patterns connected to the macrophyte
growth cycle (Figs. \ref{community} and \ref{desulfo}). Studies focusing
on functional comparisons between communities associated with different
hosts showed that the majority of functions could be found in every
community, indicating functional redundancy (Roth-Schulze \emph{et al.},
2016). This difference between phylogenetic varaibility and functional
stability was explained by the lottery hypothesis assuming an initial
random colonization step performed by a set of functionally equivalent
taxonomic groups (Burke and Peter Steinberg \emph{et al.}, 2011;
Roth-Schulze \emph{et al.}, 2016). It is possible that functional
redundancy is a characteristic of high abundance taxa detected to be
present throughout the year, while seasonal and/or host-specific
functions are an attribute of taxa displaying successional patterns.
Further studies connecting taxonomy with functional properties will be
required to elucidate the degree of functional redundancy or specificity
in epiphytic microbial communities.

@@

Illumina MiSeq sequencing of the V4 region of the 16S rRNA gene was
performed as described previously (Korlević \emph{et al.}, submitted).
The V4 region of the 16S rRNA gene was amplified using a two-step PCR
procedure. In the first PCR, the 515F (5'-GTGYCAGCMGCCGCGGTAA-3') and
806R (5'-GGACTACNVGGGTWTCTAAT-3') primers from the Earth Microbiome
Project
(\url{http://press.igsb.anl.gov/earthmicrobiome/protocols-and-standards/16s/})
were used (Caporaso \emph{et al.}, 2012; Apprill \emph{et al.}, 2015;
Parada \emph{et al.}, 2016). These primers contained on their 5' end a
tagged sequence. Purified PCR products were sent for Illumina MiSeq
sequencing at IMGM Laboratories, Martinsried, Germany. Prior to
sequencing at IMGM, the second PCR amplification of the two-step PCR
procedure was performed using primers targeting the tagged region
incorporated in the first PCR. In addition, these primers contained
adapter and sample-specific index sequences. Beside samples, a positive
and negative control for each sequencing batch was sequenced. The
negative control comprised PCR reactions without DNA template, while for
a positive control a mock community composed of evenly mixed DNA
material originating from 20 bacterial strains (ATCC MSA-1002, ATCC,
USA) was used. Sequences obtained in this study have been deposited in
the European Nucleotide Archive (ENA) at EMBL-EBI under accession number
PRJEB37267.

@@

Obtained sequences were analyzed on the computer cluster Isabella
(University Computing Center, University of Zagreb) using mothur
(version 1.43.0) (Schloss \emph{et al.}, 2009) according to the MiSeq
Standard Operating Procedure (MiSeq SOP;
\url{https://mothur.org/wiki/MiSeq_SOP}) (Kozich \emph{et al.}, 2013)
and recommendations provided by the Riffomonas project to enhance data
reproducibility (\url{http://www.riffomonas.org/}). For alignment and
classification of sequences the SILVA SSU Ref NR 99 database (release
138; \url{http://www.arb-silva.de}) was used (Quast \emph{et al.}, 2013;
Yilmaz \emph{et al.}, 2014). Pipeline data processing and visualization
was done using R (version 3.6.0) (R Core Team, 2019) combined with
packages vegan (version 2.5-6) (Oksanen \emph{et al.}, 2019), tidyverse
(version 1.3.0) (Wickham \emph{et al.}, 2019) and multiple other
packages (Xie, 2014, 2015, 2019, 2020; Neuwirth, 2014; Xie \emph{et
al.}, 2018; Wilke, 2018; Allaire \emph{et al.}, 2019; Zhu, 2019). The
detailed analysis procedure including the R Markdown file are available
in the GitHub repository
(\url{https://github.com/MicrobesRovinj/Korlevic_EpiphyticDynamics_EnvironMicrobiol_2020}).
Based on the ATCC MSA-1002 mock community included in the analysis an
average sequencing error rate of 0.01 \si{\percent} was determined,
which is in line with previously reported values for next-generation
sequencing data (Kozich \emph{et al.}, 2013; Schloss \emph{et al.},
2016). In addition, the negative controls processed together with the
samples yielded on average only 2 sequences after sequence quality
curation.

>>>>

\newpage
\linenumbers
\sisetup{mode=text}
\setlength\parindent{24pt}

@@

Marine macrophytes (seagrasses and macroalgae) are important ecosystem
engineers forming close associations with microorganisms belonging to
all three domains of life (Egan \emph{et al.}, 2013; Tarquinio \emph{et
al.}, 2019). Microbes can live within macrophyte tissue as endophytes or
can form epiphytic communities on surfaces of leaves and thalli (Egan
\emph{et al.}, 2013; Hollants \emph{et al.}, 2013; Tarquinio \emph{et
al.}, 2019). Epiphytic and endophytic microbial communities exhibit a
close functional relationship with the macrophyte host. It was proposed
that this close relationship constitutes a holobiont, an integrated
community where the macrophyte organism and its symbiotic partners
support each other (Margulis, 1991; Egan \emph{et al.}, 2013; Tarquinio
\emph{et al.}, 2019).

@@

The majority of studies describing macrophyte epiphytic microbial
communities did not include possible seasonal changes (Crump and Koch,
2008; Lachnit \emph{et al.}, 2009; Burke and Thomas \emph{et al.}, 2011;
Roth-Schulze \emph{et al.}, 2016; Ugarelli \emph{et al.}, 2019). If
seasonal changes were taken into account, low temporal frequency,
applied methodologies and/or limited number of analysed host species did
not allow a high taxonomic resolution (Tujula \emph{et al.}, 2010;
Lachnit \emph{et al.}, 2011; Miranda \emph{et al.}, 2013; Michelou
\emph{et al.}, 2013; Mancuso \emph{et al.}, 2016). In the present study
we describe the seasonal dynamics of bacterial and archaeal communities
on the surfaces of the seagrass \emph{Cymodocea nodosa} and siphonous
macroalgae \emph{Caulerpa cylindracea} determined on a mostly monthly
scale. Bacterial and archaeal epiphytes were sampled in a meadow of
\emph{C. nodosa} invaded by the invasive \emph{C. cylindracea} and in a
locality of only \emph{C. cylindracea} located in the proximity of the
seagrass meadow. For comparison, the microbial community of the ambient
seawater was also characterized.

@@

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of sequences without
known relatives (no relative sequences) and eukaryotic, chloroplast and
mitochondrial sequences (\autoref{supp-nseq_notus}). A total of 35
samples originating from epiphytic archaeal and bacterial communities
associated with surfaces of the seagrass \emph{C. nodosa} and the
macroalga \emph{C. cylindracea} were analysed. In addition, 18 samples
(one of the samples was sequenced twice) originating from the ambient
seawater were also processed for comparison. The number of reads per
sample ranged between 8,408 and 77,463 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off as commonly observed in
high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences as mentioned above reads were clustered into 28,750
different OTUs at a similarity level of 97 \si{\percent}. Read numbers
were normalized to the minimum number of sequences (8,408,
\autoref{supp-nseq_notus}) through rarefaction resulting in 17,079
different OTUs with 384 to 2,062 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes in richness
and diversity the observed number of OTUs, Chao1, ACE, Exponential
Shannon and Inverse Simpson (Jost, 2006) were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{C. cylindracea} (mixed {[}Number of OTUs, 1,684.1 ±
133.8 OTUs{]} and monospecific {[}Number of OTUs, 1,731.9 ± 163.8
OTUs{]}) than for \emph{C. nodosa} (Number of OTUs, 1,058.2 ± 214.5
OTUs) and lowest values were obtained for the microbial community of the
ambient seawater (Number of OTUs, 523.9 ± 138.7 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not reveal such large
dissimilarities. \emph{C. nodosa} communities showed a slow increase
towards the end of the study, while \emph{C. cylindracea} (mixed and
monospecific) communities were characterized by slightly higher values
in spring and summer than in autumn and winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient, respectively, were calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from the
particular environment. The highest proportion of shared OTUs and
community was found between mixed and monospecific \emph{C. cylindracea}
(Jaccard, 0.35; Bray-Curtis, 0.78), while lower shared values were
calculated between ambient seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{C. nodosa} and
\emph{C. cylindracea} were approximately in-between the values of mixed
and monospecific \emph{C. cylindracea}. To assess seasonal changes in
the proportion of shared OTUs and communities the Jaccard's and
Bray-Curtis Similarity Coefficients were calculated between consecutive
sampling points (\autoref{shared}). Both coefficients showed similar
trends. Temporal proportional changes were more pronounced for ambient
seawater than for \emph{C. nodosa} and especially \emph{C. cylindracea}
associated communities (\autoref{shared}). In addition, only 0.4 -- 1.0
\si{\percent} of OTUs from each surface associated community were
present at all seasons. These persistent OTUs constituted a high
proportion of total sequences (39.5 -- 52.3 \si{\percent}). To further
disentangle the environmental and seasonal community dissimilarity a
Principal Coordinates Analysis (PCoA) based on Bray-Curtis distances and
OTU abundances was applied. A clear separation between ambient seawater
and surface associated communities was found (\autoref{pcoa}). In
addition, a separation of epiphytic bacterial and archaeal communities
based on host species was detected. This separation was further
supported by ANOSIM (R = 0.96, \emph{p} \textless{} 0.001). Seasonal
changes of \emph{C. nodosa} associated communities indicated a
separation between spring, summer and autumn/winter samples (ANOSIM, R =
0.54, \emph{p} \textless{} 0.01). For \emph{C. cylindracea} associated
communities a separation between summer and autumn/winter/spring samples
was observed that was, however, not as strong as for \emph{C. nodosa}
associated communities (ANOSIM, R = 0.30, \emph{p} \textless{} 0.05)
(\autoref{pcoa}).

@@

We observed a strong differentiation between the surface attached and
ambient seawater communities at the level of OTUs, in agreement with
most published studies (Burke and Thomas \emph{et al.}, 2011; Michelou
\emph{et al.}, 2013; Roth-Schulze \emph{et al.}, 2016; Mancuso \emph{et
al.}, 2016; Crump \emph{et al.}, 2018; Ugarelli \emph{et al.}, 2019).
This indicates that marine macrophytes are a selecting factor from the
pool of microbial taxa present in the ambient seawater, modifying the
microbial community once the macrophyte associated microbial biofilm
develops (Salaün \emph{et al.}, 2012; Michelou \emph{et al.}, 2013). In
contrast, Fahimipour \emph{et al.} (2017) report in a global study of
\emph{Zostera marina}, similarities between the microbial community
developed on leaves and in the ambient seawater. The discrepancy between
our data and the study of Fahimipour \emph{et al.} (2017) could be
explained by different seagrass species, methodological variations or
biogeographic trends as Fahimipour \emph{et al.} (2017) analysed samples
from different locations throughout the northern hemisphere. It is
possible that the microbial communities in ambient seawater and on
leaves from the same location are differing but are still more similar
to each other when compared to other sampling locations. Indeed, it was
found that prokaryotic communities vary substantially between different
sampling sites (Bengtsson \emph{et al.}, 2017). When the taxonomic
composition at high ranks was analysed no such strong differentiation
was noticed. Phyla and classes such as \emph{Actinobacteriota},
\emph{Bacteroidota}, \emph{Cyanobacteria}, \emph{Alphaproteobacteria},
\emph{Gammaproteobacteria} and \emph{Verrucomicrobiota} were found in
both ambient seawater as well as macrophyte associated, in agreement
with previous studies (Burke and Thomas \emph{et al.}, 2011; Egan
\emph{et al.}, 2013; Michelou \emph{et al.}, 2013). In contrast, when
low taxonomic ranks were analysed (i.e., family and genus) a strong
differentiation was observed (Figs. \ref{cyano} -- \ref{desulfo}). A
similar differentiation at lower microbial taxonomic ranks between
ambient seawater and macrophytes was described for other macrophyte
species as well (Egan \emph{et al.}, 2013; Michelou \emph{et al.}, 2013;
Ugarelli \emph{et al.}, 2019).

@@

Seasonal changes in richness in the epiphytic community were substantial
as indicated by the proportion of OTUs (\(\leq\) 1.0 \si{\percent})
present at every sampling date. These persistent OTUs, however, were
accounting for a high proportion of sequences (\(\geq\) 39.5
\si{\percent}) (\autoref{shared}). A very similar proportion of
persistent OTUs was reported in high-frequency sampling studies
describing seasonal changes in picoplankton (Gilbert \emph{et al.},
2009, 2012). In comparison to the seawater community, a lower degree of
seasonal shifts was observed for the macrophyte surface associated
communities. It appears that microniches at the surfaces of macrophytes
are providing more stable conditions than the ambient seawater. At the
level of OTUs seasonal changes of \emph{C. nodosa} and \emph{C.
cylindracea} associated communities were identified that could be linked
to the growth cycle of the seagrass and macroalgae (M. Najdek, personal
communication). \emph{C. nodosa} was characterized by a spring community
during maximum seagrass proliferation, a summer community during the
highest standing stock of \emph{C.nodosa} and an autumn/winter community
during the decay of seagrass biomass. In contrast, \emph{C. cylindracea}
started to proliferate in late spring and was characterized only by a
summer community during high growth rates and by an autumn/winter/spring
community when the biomass was at the peak and decaying thereafter.
Similar seasonal changes in the epiphytic community were also described
for other macroalgae (Tujula \emph{et al.}, 2010; Lachnit \emph{et al.},
2011). Higher seasonal stability of \emph{C. cylindracea} surface
communities than in \emph{C. nodosa} was also observed in the higher
proportion of shared communities between two consecutive sampling dates
in \emph{C. cylindracea}.

@@

surface for non-specific colonizers (Zavodnik \emph{et al.}, 1998).

@@

not exhibit strong successional patterns. In every analysed group, with

@@

and no relative \emph{Desulfocapsaceae}, known sulphate sediment groups

@@

In temperate zones, marine macrophytes are exhibiting growth cycles, so
it is not surprising that the associated epiphytic microbial community
is undergoing partial seasonal changes. In the present study, we could
identify in every analysed high rank taxa phylogenetic groups present
throughout the year, comprising most of the sequences and a lower
proportion of taxa showing seasonal patterns connected to the macrophyte
growth cycle (Figs. \ref{community} and \ref{desulfo}). Studies focusing
on functional comparisons between communities associated with different
hosts showed that the majority of functions could be found in every
community, indicating functional redundancy (Roth-Schulze \emph{et al.},
2016). This difference between phylogenetic varaibility and functional
stability was explained by the lottery hypothesis assuming an initial
random colonization step performed by a set of functionally equivalent
taxonomic groups (Burke and Peter Steinberg \emph{et al.}, 2011;
Roth-Schulze \emph{et al.}, 2016). It is possible that functional
redundancy is a characteristic of high abundance taxa detected to be
present throughout the year, while seasonal and/or host-specific
functions are an attribute of taxa displaying successional patterns.
Further studies connecting taxonomy with functional properties will be
required to elucidate the degree of functional redundancy or specificity
in epiphytic microbial communities.

@@

Illumina MiSeq sequencing of the V4 region of the 16S rRNA gene was
performed as described previously (Korlević \emph{et al.}, submitted).
The V4 region of the 16S rRNA gene was amplified using a two-step PCR
procedure. In the first PCR, the 515F
(\(5'\)-GTGYCAGCMGCCGCGGTAA-\(3'\)) and 806R
(\(5'\)-GGACTACNVGGGTWTCTAAT-\(3'\)) primers from the Earth Microbiome
Project
(\url{http://press.igsb.anl.gov/earthmicrobiome/protocols-and-standards/16s/})
were used (Caporaso \emph{et al.}, 2012; Apprill \emph{et al.}, 2015;
Parada \emph{et al.}, 2016). These primers contained on their \(5'\) end
a tagged sequence. Purified PCR products were sent for Illumina MiSeq
sequencing at IMGM Laboratories, Martinsried, Germany. Prior to
sequencing at IMGM, the second PCR amplification of the two-step PCR
procedure was performed using primers targeting the tagged region
incorporated in the first PCR. In addition, these primers contained
adapter and sample-specific index sequences. Beside samples, a positive
and negative control for each sequencing batch was sequenced. The
negative control comprised PCR reactions without DNA template, while for
a positive control a mock community composed of evenly mixed DNA
material originating from 20 bacterial strains (ATCC MSA-1002, ATCC,
USA) was used. Sequences obtained in this study have been deposited in
the European Nucleotide Archive (ENA) at EMBL-EBI under accession number
PRJEB37267.

@@

Obtained sequences were analysed on the computer cluster Isabella
(University Computing Center, University of Zagreb) using mothur
(version 1.43.0) (Schloss \emph{et al.}, 2009) according to the MiSeq
Standard Operating Procedure (MiSeq SOP;
\url{https://mothur.org/wiki/MiSeq_SOP}) (Kozich \emph{et al.}, 2013)
and recommendations provided by the Riffomonas project to enhance data
reproducibility (\url{http://www.riffomonas.org/}). For alignment and
classification of sequences the SILVA SSU Ref NR 99 database (release
138; \url{http://www.arb-silva.de}) was used (Quast \emph{et al.}, 2013;
Yilmaz \emph{et al.}, 2014). Pipeline data processing and visualization
was done using R (version 3.6.0) (R Core Team, 2019) combined with
packages vegan (version 2.5-6) (Oksanen \emph{et al.}, 2019), tidyverse
(version 1.3.0) (Wickham \emph{et al.}, 2019) and multiple other
packages (Xie, 2014, 2015, 2019, 2020; Neuwirth, 2014; Xie \emph{et
al.}, 2018; Wilke, 2018; Allaire \emph{et al.}, 2019; Zhu, 2019). The
detailed analysis procedure including the R Markdown file are available
in the GitHub repository
(\url{https://github.com/MicrobesRovinj/Korlevic_EpiphyticDynamics_EnvironMicrobiol_2020}).
Based on the ATCC MSA-1002 mock community included in the analysis an
average sequencing error rate of 0.01 \si{\percent} was determined,
which is in line with previously reported values for next-generation
sequencing data (Kozich \emph{et al.}, 2013; Schloss \emph{et al.},
2016). In addition, the negative controls processed together with the
samples yielded on average only 2 sequences after sequence quality
curation.
